---
title: "An alternative model for diagnosis probability"
author: "Rachael Caelie (Rocky) Aikens"
date: "7/14/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, cache = TRUE, fig.align = "center", fig.height = 3)
library(confirmationBias)
library(tidyverse)
library(ggpubr)
library(knitr)
library(binom)

theme_set(theme_light())

n <- 50000
```

# Setup

## A model for Longitudinal Disease Progression

Now let's imagine we're dealing with a disease that begins at a low severity and progresses every year.  We'll assume that severity progresses from 0 to 1 according to a sigmoid function with the following parameters:

$$R_i(t) = \frac{1}{1 + e^{-\beta_i(t - T_i)}}.$$
Here, $\beta_i$ and $T_i$ are intrinsic to the individual:

\begin{itemize}
  \item $\beta_i$ is the slope of progression. A larger $\beta_i$ indicates a quickly progressing disease.
  \item $T_i$ controlls the time of onset.  $T_i$ is the time when severity reaches $0.5$
\end{itemize}

We'll simulate $\beta_i$ and $T_i$ from the following distributions:

\begin{align*}
\beta_i &\sim_{iid} \text{Beta}(\alpha = 2, \beta = 6)\\
T_i &\sim_{iid} \text{Normal}(\mu = 50, \sigma = 5)
\end{align*}

## Model: Diagnosis rate depends on severity and baseline covariates

Now, let's suppose that the probability of diagnosis depends on severity and baseline characteristics:

$$\phi(R_i, X_i) = \frac{1 + \beta_2X_i}{1 + exp(-(\beta_0 + \beta_1R_i))} - c_{x_i},$$

where again, $c_{x_i} = \frac{1}{1 + exp(-(\beta_0 + \beta_2X_i))}$ is a corrective constant to ensure that $\phi(0, X_i) = 0$. Also, let $\beta_2 = -\frac{1}{2}$.

Inessence, this means that people with $X_i = 0$ are diagnosed in the same way as above, but people with $X_i = 1$ are diagnosed with a lower probability, shown below:

```{r}
R <- seq(0, 1, length.out = 10000)

plt_data <- expand.grid(R, c(0,1)) %>%
  rename(R = Var1, X = Var2) %>%
  mutate(phi = (1 - 0.5* X)/(1 + exp(-(20*R - 2.5*X - 10)))) %>%
  mutate(X = as.factor(X))

ggplot(data = plt_data, aes(x = R, y = phi, group = X, color = X)) +
  geom_line() + xlim(c(0,1)) + ylim(c(0,1)) + coord_fixed() +
  scale_color_brewer(palette = "Set1")
```

An example of how this plays out is shown below on a subset of the data.  People with $X = 0$ tend to be diagnosed earlier in their disease progression than people with $X = 1$.

```{r}
set.seed(123)
eg_long <- generate_longitudinal(n = 20)
```

```{r}
eg_diagnosed <- diagnose_longitudinal(eg_long)

diagnosis_pts <- eg_diagnosed %>%
  group_by(id) %>%
  arrange(-diagnosed, t) %>%
  slice_head(1)

ggplot(eg_diagnosed, aes(x = t, y = severity, group = id, color = as.factor(x), alpha = diagnosed)) +
  geom_line() +
  geom_point(data = diagnosis_pts, mapping = aes(x = t, y = severity)) +
  scale_color_brewer(palette = "Set1")
```

```{r}
eg_diagnosed %>%
  group_by(id) %>%
  mutate(cum_p_diagnosed = 1 - cumprod(1 - p_diagnose)) %>%
  ggplot(aes(x = t, y = cum_p_diagnosed, group = id, color = as.factor(x))) +
  geom_line() +
  scale_color_brewer(palette = "Set1") +
  ylab("Probability of diagnosis by time t")
```

Under this model (where diagnosis probability depends on X), let's see what we glean from a naive study of these data.

```{r}
n <- 5000
```

Let's suppose there are `r n` individuals with this disease in the EHR.  We don't observe their underlying disease state; just their diagnosis.  We also don't observe the severity of their illness prior to diagnosis.

```{r generate data}
set.seed(123)
long_df <- generate_longitudinal(n = n)
```

```{r}
full_long_df <- diagnose_longitudinal(long_df)

observed_long_df <- full_long_df %>%
  filter(diagnosed == T)

observed_short_df <- observed_long_df %>%
  group_by(id) %>%
  arrange(-diagnosed, t) %>%
  slice_head(1)

# same as observed_short_df when every individual is diagnosed
full_short_df <- full_long_df %>%
  group_by(id) %>%
  arrange(-diagnosed, t) %>%
  slice_head(1)
```

```{r}
observed_short_df %>%
  group_by(x) %>%
  summarize(n = n())
full_short_df %>%
  group_by(x) %>%
  summarize(n = n())
```

In this set-up, `r dim(observed_short_df)[1]` individuals out of `r n` with the disease are diagnosed.  All of the undiagnosed individuals have $X = 1$. The plot below shows the severity at onset for individuals with $x = 0$ compared to $x = 1$.  This probably is indicative of the fact that the people with $x = 0$ tend to be diagnosed at the start of their disease onset, while the people with $x = 1$ tend to be diagnosed later in their disease progression.

```{r}
ggplot(mutate(observed_short_df, x = as.factor(x)),
       aes(x = severity, y=..density.., group = x, fill = x)) +
  geom_histogram(position = "identity", alpha = 0.6, binwidth = 0.01) +
  scale_fill_brewer(palette = "Set1")
```
