---
title: "EBM with Doctor Subtypes"
author: "Rachael Caelie (Rocky) Aikens"
date: "8/3/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
library(confirmationBias)
library(tidyverse)
library(ggpubr)
library(knitr)
library(binom)

theme_set(theme_light())

n <- 5000
prev <- 0.15
```

# Doctor variation

Now let's take a doctor-centric view.  Suppose that doctors (indexed by $j$) diagnose with a probability that depends on (1) the representativeness of the patient's symptoms, $R_i$ (2) the patients demographics $X_i$, and (3) the parameters of their own diagnostic habits, $\boldsymbol{\theta}_j = (\beta_{0j}, \beta_{1j}, \beta_{2j})$. Explicitly, for patient $i$ seeing doctor $j$, 

$$D_i \sim_{iid} \text{Bernoulli}\left(\phi(R_i, X_i, \boldsymbol{\theta}_j)\right)$$
where, 

$$\phi(R_i, X_i) = \frac{1}{1 + exp(-(\beta_{0j} + \beta_{1j}R_i + \beta_{2j}X_i))} - c_{x_i \boldsymbol{\theta_j}}.$$

## Two types of doctors:

Recall that an ideal diagnostic model would have $\beta_1 = \infty$ and $\beta_2 = 0$, since the only feature that really matters is the symptoms, and in this world the symptoms only emerge when the disease is present.  However, doctors aren't perfect at recognizing the symptoms of this disease.  Instead, our doctors have two patterns:

\begin{align*}
\boldsymbol{\theta_A} &= (-10, 20, 0)
\boldsymbol{\theta_B} &= (-10, 20, -10)
\end{align*}

In other words, "A"-type doctors diagnose people based on the representativeness of their symptoms only, whereas "B"-type doctors diagnose $X_i = 0$ people at the same rate as A-doctors but diagnose $X_i = 1$ people at a lower rate.  In our case, we know A-doctors will have a lower misdiagnosis rate than B-doctors.

## A simple doctor population

Now let's imagine we have a population of 100 A-doctors and 100 B-doctors.  Each doctor sees 200 patients a year. A histogram rates of correct diagnosis of disease patients for each doctor is shown below, first as an aggregate, then subsetted by X.  

```{r setup params}
n_A_docs <- 100
n_B_docs <- 100
theta_A <- c(-10, 20, 0)
theta_B <- c(-10, 20, -10)

patients_per_doc <- 300
```


```{r}
df_A <- generate_cross_sectional(n = patients_per_doc*n_A_docs, prevalence = 0.15) %>%
  mutate(doctor_id = rep(1:n_A_docs, each = patients_per_doc),
         doctor_type = "A") %>%
  diagnose_cross_sectional(theta = theta_A)

df_B <- generate_cross_sectional(n = patients_per_doc*n_B_docs, prevalence = 0.15) %>%
  mutate(doctor_id = rep((n_A_docs + 1):(n_A_docs + n_B_docs), each = patients_per_doc),
         doctor_type = "B") %>%
  diagnose_cross_sectional(theta = theta_B)

full_df <- rbind(df_A, df_B)
```

```{r}
full_df %>% group_by(doctor_id, doctor_type, x) %>%
  filter(disease == T) %>%
  summarize(correct_diagnosis_rate = mean(diagnosed == T)) %>%
  ggplot(aes(x = correct_diagnosis_rate, y = ..density.., group = doctor_type, fill = doctor_type)) +
  geom_histogram(position = "identity", alpha = 0.5) +
  scale_fill_brewer(palette = "Set2", name = "Doctor Type") 

full_df %>% group_by(doctor_id, doctor_type, x) %>%
  filter(disease == T) %>%
  summarize(correct_diagnosis_rate = mean(diagnosed == T)) %>%
  ggplot(aes(x = correct_diagnosis_rate, y = ..density.., group = doctor_type, fill = doctor_type)) +
  geom_histogram(position = "identity", alpha = 0.5) +
  scale_fill_brewer(palette = "Set2", name = "Doctor Type") +
  facet_wrap(~x) 
```

As expected, the doctors have similar diagnosis patterns when $X = 0$, but type B doctors perform much more poorly when patients with $X = 1$ are considered.

## A data-driven diagnostic model

```{r}
glm(diagnosed ~ x + representativeness, data = rbind(df_A, df_B), family = "binomial")
```

# What happens over time

```{r}
simulate_EBM_docs_vary <- function(years = 10, evidence_rate = 0.1){
  for (i in 1:years){
    # generate undiagnosed data
    curr_df_A <- df_A <- generate_cross_sectional(n = patients_per_doc*n_A_docs, prevalence = 0.15) %>%
      mutate(doctor_id = rep(1:n_A_docs, each = patients_per_doc),
         doctor_type = "A") %>%
      mutate(study_id = i - 1)
    
    curr_df_B <- generate_cross_sectional(n = patients_per_doc*n_B_docs, prevalence = 0.15) %>%
      mutate(doctor_id = rep((n_A_docs + 1):(n_A_docs + n_B_docs), each = patients_per_doc),
         doctor_type = "B") %>%
      mutate(study_id = i - 1 )
  
    if (i == 1){
      # first year, let docs diagnose based on own preferences
      curr_df_A <- diagnose_cross_sectional(curr_df_A, theta = theta_A) %>%
        mutate(p_model = NA, p_doc = p_diagnose) %>%
        select(x, disease, representativeness, doctor_id, doctor_type, study_id, p_model, p_doc, p_diagnose, diagnosed)
      
      full_df_A <- curr_df_A
      
      curr_df_B <- diagnose_cross_sectional(curr_df_B, theta = theta_B) %>%
        mutate(p_model = NA, p_doc = p_diagnose) %>%
        select(x, disease, representativeness, doctor_id, doctor_type, study_id, p_model, p_doc, p_diagnose, diagnosed)
      
      full_df_B <- curr_df_B
    } else{
      # subsequent years, docs increasingly diagnose based on model
      evidence_weight <- (i - 1)*evidence_rate
      
      nrows_A <- dim(curr_df_A)[1]
      nrows_B <- dim(curr_df_B)[1]
      
      # diagnose patients
      curr_df_A <- curr_df_A %>%
        mutate(p_model = predict(curr_model, newdata = curr_df_A, type = "response")) %>%
        mutate(p_doc = diagnosis_fn(representativeness, x, theta = theta_A)) %>%
        mutate(p_diagnose = (1 - evidence_weight)*p_doc + evidence_weight*p_model) %>%
        mutate(diagnosed = rbinom(nrows_A, size = 1, prob = p_diagnose))
      
      curr_df_B <- curr_df_B %>% 
        mutate(p_model = predict(curr_model, newdata = curr_df_B, type = "response")) %>%
        mutate(p_doc = diagnosis_fn(representativeness, x, theta = theta_B)) %>%
        mutate(p_diagnose = (1 - evidence_weight)*p_doc + evidence_weight*p_model) %>%
        mutate(diagnosed = rbinom(nrows_B, size = 1, prob = p_diagnose))
      
      # store results
      full_df_A <- rbind(full_df_A, curr_df_A)
      full_df_B <- rbind(full_df_B, curr_df_B)
    }
  # build EBM model based on this year's diagnosis
  curr_model <- glm(diagnosed ~ x + representativeness, data = rbind(curr_df_A, curr_df_B), family = "binomial")
  }
  return(rbind(full_df_A, full_df_B))
}
```

# Homogenous Doctors

First, let's suppose all doctors have the same diagnostic habits.  Let $\theta_A = \theta_B = (-10, 20, -5)$

```{r}
n_A_docs <- 25
n_B_docs <- 25
theta_A <- c(-10, 20, -5)
theta_B <- c(-10, 20, -5)

patients_per_doc <- 300

set.seed(123)
time_df <- simulate_EBM_docs_vary(years = 10)
```

```{r}
plt_data <- time_df %>%
  filter(disease == TRUE) %>%
  group_by(x, study_id, doctor_id, doctor_type) %>%
  summarize(correct_diagnosis_rate = mean(diagnosed == 1)) %>%
  mutate(x = as.factor(x))

ggplot(plt_data, aes(x = study_id, y = correct_diagnosis_rate, group = doctor_id)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~x) + 
  scale_x_continuous(name = "Years of Study", breaks = 0:9) +
  scale_y_continuous(name = "Correct Diagnosis Rate",
                     limits = c(0,0.9),
                     breaks = seq(0, 0.9, by = 0.1),
                     labels = scales::percent_format(accuracy = 1))
```



# Variable Doctors

Now we imagine that this is part of a learning healthcare system.  Every year, a new batch of patients comes to the hospital, and is diagnosed (or isn't).  Every year, the researchers retrain their model based on the previous year's diagnoses.

```{r}
n_A_docs <- 25
n_B_docs <- 25
theta_A <- c(-10, 20, 0)
theta_B <- c(-10, 20, -10)

patients_per_doc <- 300

set.seed(123)
time_df <- simulate_EBM_docs_vary(years = 10)
```

```{r}
plt_data <- time_df %>%
  filter(disease == TRUE) %>%
  group_by(x, study_id, doctor_id, doctor_type) %>%
  summarize(correct_diagnosis_rate = mean(diagnosed == 1)) %>%
  mutate(x = as.factor(x))

ggplot(plt_data, aes(x = study_id, y = correct_diagnosis_rate, group = doctor_id, color = doctor_type)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~x) + 
  scale_color_brewer(palette = "Set2", name = "Doctor Type") +
  scale_x_continuous(name = "Years of Study", breaks = 0:9) +
  scale_y_continuous(name = "Correct Diagnosis Rate",
                     limits = c(0,0.9),
                     breaks = seq(0, 0.9, by = 0.1),
                     labels = scales::percent_format(accuracy = 1))
```


