---
title: "EBM with Doctor Subtypes"
author: "Rachael Caelie (Rocky) Aikens"
date: "8/3/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
library(confirmationBias)
library(tidyverse)
library(ggpubr)
library(knitr)
library(binom)

theme_set(theme_light())

n <- 5000
prev <- 0.15
```

# Doctor variation

Now let's take a doctor-centric view.  Suppose that doctors (indexed by $j$) diagnose with a probability that depends on (1) the severity of the patient's symptoms, $S_i$ (2) the patients demographics $X_i$, and (3) the parameters of their own diagnostic habits, $\boldsymbol{\theta}_j = (\beta_{0j}, \beta_{1j}, \beta_{2j})$. Explicitly, for patient $i$ seeing doctor $j$, 

$$D_i \sim_{iid} \text{Bernoulli}\left(\phi(S_i, X_i, \boldsymbol{\theta}_j)\right)$$
where, 

$$\phi(S_i, X_i) = \frac{1}{1 + exp(-(\beta_{0j} + \beta_{1j}S_i + \beta_{2j}X_i))} - c_{x_i \boldsymbol{\theta_j}}.$$

## Doctors vary:

Recall that an ideal diagnostic model would have $\beta_1 = \infty$ and $\beta_2 = 0$, since the only feature that really matters is the symptoms, and in this world the symptoms only emerge when the disease is present.  However, doctors aren't perfect at recognizing the symptoms of this disease.  Instead, our doctors have their own diagnostic preferences. Some weight the symptoms more heavily and largely ignore demographics; others weight the demographics heavily and are unlikely to diagnose most cases with the disease in the underdiagnosed demographic group.

## A simple doctor population

Now let's imagine we have a population of 200 doctors with their own diagnostic practices.  Each doctor sees 200 patients a year. A histogram rates of correct diagnosis of disease patients for each doctor is shown below, first as an aggregate, then subsetted by X.

```{r setup params}
n_docs <- 200
patients_per_doc <- 1000
```


```{r}
df <- generate_cs_with_doctors(patients_per_doc, n_docs, sigma = 5) %>%
  diagnose_cs_with_doctors()

doc_params <- df %>% 
  group_by(doctor_id) %>% 
  summarize(beta0 = first(beta0), betaS = first(betaS), betaX = first(betaX))
```

```{r}
S <- seq(0, 1, length.out = 100)
sample_docs <- expand.grid(S, c(0,1), c(1:5)) %>%
  rename(S = Var1, X = Var2, doctor_id = Var3) %>%
  left_join(doc_params, by = "doctor_id") %>%
  mutate(p_diagnose = diagnosis_fn_by_betas(S, X, beta0, betaS, betaX))

ggplot(sample_docs, aes(x = S, y = p_diagnose, group = doctor_id, color = as.factor(X))) +
  geom_line() + 
  facet_wrap(~X) +
  scale_color_brewer(palette = "Set1")
```


```{r}
df %>% group_by(doctor_id, x) %>%
  filter(disease == T) %>%
  summarize(correct_diagnosis_rate = mean(diagnosed == T)) %>%
  ggplot(aes(x = correct_diagnosis_rate, y = ..density..)) +
  geom_histogram(position = "identity", alpha = 0.5) 

df %>% group_by(doctor_id, x) %>%
  filter(disease == T) %>%
  summarize(correct_diagnosis_rate = mean(diagnosed == T)) %>%
  ggplot(aes(x = correct_diagnosis_rate, y = ..density..)) +
  geom_histogram(position = "identity", alpha = 0.5) +
  facet_wrap(~x) 

df %>% group_by(doctor_id, x) %>%
  filter(disease == T) %>%
  summarize(correct_diagnosis_rate = mean(diagnosed == T),
            beta0 = first(beta0),
            betaS = first(betaS),
            betaX = first(betaX)) %>%
  ggplot(aes(x = betaX, y = correct_diagnosis_rate, color = as.factor(x))) +
  geom_point() +
  scale_color_brewer(palette = "Set1") +
  facet_wrap(~x)

df %>% group_by(doctor_id) %>%
  filter(disease == T) %>%
  summarize(correct_diagnosis_rate = mean(diagnosed == T),
            beta0 = first(beta0),
            betaS = first(betaS),
            betaX = first(betaX)) %>%
  ggplot(aes(x = betaX, y = correct_diagnosis_rate)) +
  geom_point()

```

As expected, the doctors have similar diagnosis patterns when $X = 0$, but type B doctors perform much more poorly when patients with $X = 1$ are considered.

## A data-driven diagnostic model

```{r}
glm(diagnosed ~ x + severity, data = df, family = "binomial")
```

# What happens over time

```{r}
simulate_EBM_docs_vary <- function(sigma, n_docs, patients_per_doc = 300,
                                   years = 10, evidence_rate = 0.1){
  for (i in 1:years){
    # generate undiagnosed data
    curr_df <- generate_cs_with_doctors(patients_per_doc, n_docs, sigma = sigma) %>%
      mutate(study_id = i - 1)
  
    if (i == 1){
      # first year, let docs diagnose based on own preferences
      curr_df <- diagnose_cs_with_doctors(curr_df) %>%
        mutate(p_model = NA, p_doc = p_diagnose) %>%
        select(x, disease, severity, doctor_id, beta0, betaS, betaX,
               study_id, p_model, p_doc, p_diagnose, diagnosed)
      
      full_df <- curr_df
      
    } else{
      # subsequent years, docs increasingly diagnose based on model
      # TODO: make this something more realistic/flexible
      evidence_weight <- (i - 1)*evidence_rate
      
      nrows <- dim(curr_df)[1]

      # diagnose patients
      curr_df <- curr_df %>%
        mutate(p_model = predict(curr_model, newdata = curr_df, type = "response")) %>%
        mutate(p_doc = diagnosis_fn_by_betas(severity, x, beta0, betaS, betaX)) %>%
        mutate(p_diagnose = (1 - evidence_weight)*p_doc + evidence_weight*p_model) %>%
        mutate(diagnosed = rbinom(nrows, size = 1, prob = p_diagnose))
      
      # store results
      full_df <- rbind(full_df, curr_df)
    }
  # build EBM model based on this year's diagnosis
  curr_model <- glm(diagnosed ~ x + severity, data = curr_df, family = "binomial")
  }
  return(full_df)
}
```

# Homogenous Doctors

First, let's suppose all doctors have the same diagnostic habits.  Let $\theta_A = \theta_B = (-10, 20, -5)$

```{r}
set.seed(1243)
time_df <- simulate_EBM_docs_vary(sigma = 5, n_docs = 50, patients_per_doc = 700, years = 10)
```

```{r}
plt_data <- time_df %>%
  filter(disease == TRUE) %>%
  group_by(x, study_id, doctor_id) %>%
  summarize(correct_diagnosis_rate = mean(diagnosed == 1),
            beta0 = first(beta0), 
            betaS = first(betaS),
            betaX = first(betaX)) %>%
  mutate(x = as.factor(x))

ggplot(plt_data, aes(x = study_id, y = correct_diagnosis_rate, group = doctor_id)) +
  geom_line(alpha = 0.5) +
  facet_wrap(~x) + 
  scale_x_continuous(name = "Years of Study", breaks = 0:9) +
  scale_y_continuous(name = "Correct Diagnosis Rate",
                     limits = c(0,0.9),
                     breaks = seq(0, 0.9, by = 0.1),
                     labels = scales::percent_format(accuracy = 1)) 
```

