---
title: "Diagnosis Function"
author: "Rachael Caelie (Rocky) Aikens"
date: "6/26/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
library(confirmationBias)
library(tidyverse)
library(ggpubr)
library(knitr)
library(binom)

theme_set(theme_light())

n <- 50000
```


This is a quick and dirty space for me to play with some different diagnosis functions.

```{r}
S <- seq(0, 1, length.out = 100)
```

# Identity

One thing that I've been doing for simplicity is using the identity function

$$\phi(S_i) = S_i$$
```{r}
ggplot(data = data.frame(S), aes(x = S, y = S)) +
  geom_line() + xlim(c(0,1)) + ylim(c(0,1)) + coord_fixed()
```

# Sigmoid

An intuitive tool could be the sigmoid function:

$$\phi(S_i) = \frac{1}{1 + exp(-S_i)}$$

```{r}
ggplot(data = data.frame(S), aes(x = S, y = 1/(1 + exp(-S)))) +
  geom_line() + xlim(c(0,1)) + ylim(c(0,1)) + coord_fixed()
```

This could be parametrized a little better for the $[0,1] \rightarrow [0,1]$ mapping:

$$\phi(S_i) = \frac{1}{1 + exp(-(\beta_0 + \beta_1S_i))}$$

For example, letting $\beta_1 = 15$ and $\beta_0 = -7.5$ gives a function more like:

```{r}
ggplot(data = data.frame(S), aes(x = S, y = 1/(1 + exp(-(15*S - 7.5))))) +
  geom_line() + xlim(c(0,1)) + ylim(c(0,1)) + coord_fixed()
```

Note: Letting $\beta_0 = {\beta_1}{2}$ fixes $\phi(0.5) = 0.5$

Now let's suppose we want $\phi()$ to be a function of $X_i$. We can let

$$\phi(S_i) = \frac{1}{1 + exp(-(\beta_0 + \beta_1S_i + \beta_2X_i))}$$

Now suppose that $\beta_2 = -5$

```{r}
plt_data <- expand.grid(S, c(0,1)) %>%
  rename(S = Var1, X = Var2) %>%
  mutate(phi = 1/(1 + exp(-(15*S - 3.75 * X - 7.5)))) %>%
  mutate(X = as.factor(X))

ggplot(data = plt_data, aes(x = S, y = phi, group = X, color = X)) +
  geom_line() + xlim(c(0,1)) + ylim(c(0,1)) + coord_fixed() +
  scale_color_brewer(palette = "Set1")
```

